---
title: "PHI512 Review"
author: "Erh-Chi Hsu"
date: "12/20/2021"
output: 
  html_document:
    df_print: paged
    highlight: default
    toc: true
    toc_float: true
    theme: united
    fig_width: 7
    fig_height: 4
    fig_caption: true
    code_folding: show
---
```{r library, echo = TRUE, message = FALSE, warning = FALSE}
library(tidyverse)
library(rigr)
library(knitr)
library(dplyr)
library(epiR)
```

# Basic

The datasets' names I used in this file include wcgs, fp, aug, heart, chs_recode, carotine, mri, psa, and sep.

## Read dataset

```{r ,echo = TRUE, message = FALSE, warning = FALSE}
#read and define the file
wcgs <- read.csv("~/Google Drive/0_MPH/2021fall/2PHI512Analytic/wcgs.csv")
fp <- read.csv("~/Google Drive/0_MPH/2021fall/2PHI512Analytic/dfmo.csv")
aug <- read.csv("~/Google Drive/0_MPH/2021fall/2PHI512Analytic/augmentation.csv")
heart <- read.csv("~/Google Drive/0_MPH/2021fall/2PHI512Analytic/assignment/problemset12/heart.csv")
chs_recode <- read.csv("~/Google Drive/0_MPH/2021fall/2PHI512Analytic/assignment/problemset13/chs_recode.csv")
carotine <- read.csv("~/Google Drive/0_MPH/2021fall/2PHI512Analytic/R/lab9/carotine.csv")
#read file from a website
mri <- read.table("http://www.emersonstatistics.com/datasets/mri.txt",
                  header=TRUE)
psa <- read.table("http://www.emersonstatistics.com/datasets/psa.txt",
header=TRUE)
sep <- read.table("http://www.emersonstatistics.com/datasets/sep.txt", header = TRUE)
```

## Take a look into the data

```{r}
w.names<-names(wcgs) 
w.names #variables names
w.number<-ncol(wcgs)
w.number #number of variables
w.obs<-nrow(wcgs)
w.obs #number of observations
```

## Remove NA

```{r}
#na.rm = TRUE
test_vec <- c(1, 4, NA, 12, 21, 14)
mean(test_vec, na.rm = TRUE) 
quantile(test_vec, probs = 0.5, na.rm = TRUE)
#remove NA in a certain column (change the original dataset)
mri <- mri[!is.na(mri$ldl),]
#remove NA in a certain column (create a new variable)
mri$death5 <- NA
mri$death5[mri$obstime >= 5*365.25] <- 0 
mri$death5[mri$obstime < 5*365.25] <- 1 
mri$highldl <- NA
mri$highldl[mri$ldl < 160] <- 0
mri$highldl[mri$ldl >= 160] <- 1
#example
psa$severe <- NA
psa$severe[psa$bss < 3] <- 0
psa$severe[psa$bss == 3] <- 1
psa$psaover100 <- NA
psa$psaover100[psa$pretxpsa <= 100] <- 0
psa$psaover100[psa$pretxpsa > 100] <- 1
```

## Create dataframe

```{r}
withIVT <- c(70, 75, 86, 90, 96, 102, 110, 114, 116, 117,
120, 130, 136, 142, 145, 152, 170, 182)
withoutIVT <- c(40, 50, 65, 70, 76, 78, 80, 82, 85, 88,
88, 89, 90, 94, 95, 96, 98, 98, 100, 102,
105, 107, 109, 110, 114, 115, 120, 125,
130, 135, 138, 150, 165, 170, 180)
qrs <- data.frame(time = c(withIVT, withoutIVT), IVT = c(rep("with", length(withIVT)), rep("without", length(withoutIVT))))
qrs
```

## Create 2 by 2 table

```{r}
#margin 1 means using rows as total.
table(mri$highldl,mri$death5) %>% prop.table(margin=1)
```

## Get a certain proportion or quantity

```{r}
#the proportion below a certain point (x, mean, sd)
pnorm(88, 120, 20)
#the proportion above a certain point
1 - pnorm(128, 120, 20)
#the quantity below a certain proportion (prop, mean, sd)
qnorm(0.25, 120, 20)
```

## Getting a standard deviation and confidence interval

```{r}
mean_B <- wcgs %>% filter(dibpat == "Type B") %>% pull(sbp) %>% mean()
mean_sd <- 17.2/sqrt(1589)
mean_B + c(-1.96*mean_sd, 1.96 * mean_sd)
```


# Continuous data

## Measure of spread

```{r}
range(wcgs$age)
quantile(wcgs$age, probs = c(0.25, 0.75))
sd(wcgs$age)
var(wcgs$age)
```


## Descriptive summary

```{r}
#descriptive summary by piping. 
#Cons: no SD, bugs occur when using groupby
wcgs %>% filter(age>45) %>% pull(sbp) %>% summary()
#Pros: has N, has SD, min and max
#Cons: bugs occur when using groupby
wcgs %>% filter(age>45) %>% pull(sbp) %>% descrip()
#a more complicated way to pipe groupby summary
wcgs %>% group_by(dibpat) %>% summarize(count=n(), mean=mean(age), median=median(age), sd=sd(age), IQR=IQR(age), range=(max(age)-min(age)))
```

```{r}
#descriptive summary by tapply
#Pros: has SD, don't need to use groupby
#tapply(continuous, binary, descrip)
tapply(wcgs$sbp, wcgs$smoke, descrip)
```

```{r}
#fancy way to creat descriptive table
t1<- heart %>% group_by(txt) %>% summarise(
  'N' = length(pdi),
  'Missing' = sum(is.na(pdi)),
  'Mean' = mean(pdi, na.rm = T), 
  'SD' = sd(pdi, na.rm = T),
  'Median' = median(pdi, na.rm = T),
  'IQR' = IQR(pdi, na.rm = T), 
  'Range' = max(pdi, na.rm = T) - min(pdi, na.rm = T))  %>% round(2)

t2<- heart %>% summarise(
  'txt' = 3,
  'N' = length(pdi),
  'Missing' = sum(is.na(pdi)),
  'Mean' = mean(pdi, na.rm = T), 
  'SD' = sd(pdi, na.rm = T),
  'Median' = median(pdi, na.rm = T),
  'IQR' = IQR(pdi, na.rm = T), 
  'Range' = max(pdi, na.rm = T) - min(pdi, na.rm = T))  %>% round(2)

table1 <- rbind(t1,t2)
colnames(table1)[1] <- ''

table1[,1] <- c('Circulatory Arrest','Low-flow Bypass','Overall')
t(table1) %>% kable()
```


## Visualized descriptive data

### Boxplot

```{r fig.height = 7, fig.width = 5.5}
#boxplot method 1 
treatment0<-fp%>%filter(dfmo==0)%>%pull(put0)
treatment6<-fp%>%filter(dfmo==0)%>%pull(put6)
control0<-fp%>%filter(dfmo==1)%>%pull(put0)
control6<-fp%>%filter(dfmo==1)%>%pull(put6)
boxplot(treatment0,treatment6,control0, control6, xlab= "Treatment Status", ylab= "putrescine level (umol/mg)", main="putrescine level by groups from baseline to 6 months", names=c("no DFMO treatment","no DFMO treatment", "DFMO treatment", "DFMO treatment"),ylim=c(0,9),las=1, boxwex = 0.5,col = c("pink", "purple","pink", "purple"), data=fp)
#color indication
legend(3, 9, c("baseline", "6 months"),
       fill = c("pink", "purple"))
```

```{r}
#boxplot method 2
boxplot(fp$put0 ~factor(dfmo), xlab= "Treatment Status", ylab= "putrescine level (umol/mg)", main="putrescine level by groups from baseline (black) to 6 months (red)", names=c("no DFMO", "DFMO treatment"),data=fp)
boxplot(fp$put6 ~factor(dfmo), xlab= "Treatment Status", ylab= "putrescine level (umol/mg)",ylim=c(0,2.5), names=c("no DFMO", "DFMO treatment"),data=fp,add=TRUE, border="red")
```

### Histogram

```{r}
#pipe out a histogram
wcgs %>% filter(dibpat == "Type B") %>% pull(sbp) %>% hist(main = "Histogram of Systolic Blood Pressure for Type B Participants", xlab = "Systolic Blood Pressure (mm Hg)")
#use base R
hist(wcgs[wcgs$dibpat == "Type B",c("sbp")], main = "Histogram of Systolic Blood Pressure for Type B Participants", xlab = "Systolic Blood Pressure (mm Hg)")
```

```{r}
#create fancy overlapping histogram
null <- rbinom(1e6,200,0.66)/200
alt <- rbinom(1e6,200,0.75)/200
hist(null, breaks = 30, col = rgb(0,0,1,0.5), xlab = "Proportion of pts w/ clinical improvement", xlim = c(0.5,1), main = "Plot of sampling dist under null, and suspected alternative", freq = FALSE, ylim=c(0,20), cex.main = 0.6)
abline(v = quantile(null, 0.95), col = "black", lwd = 3)
h <- hist(alt, breaks = 30, col = rgb(1,0,0,0.5), freq = FALSE, add = TRUE)
cuts <- cut(h$breaks, c(-Inf,quantile(null, 0.95),Inf))
cutscols <- ifelse(cuts == cuts[1], rgb(1,0,0,0.5), rgb(0,1,0,0.5))
plot(h, add = TRUE, col = cutscols, freq = FALSE)
```


### Density plot

```{r}
d1 <- density(wcgs$age)
plot(d1)
d2 <- density(wcgs$age, bw = 5)
plot(d2, xlab = "Age", main = "Distribution of Age")
```

### Scatterplot

```{r}
plot(sbp ~ weight, data = wcgs, 
     xlab = "Weight (lbs)", ylab = "Systolic blood pressure (mm Hg)",
     pch = 16, cex = 0.6)
```




# Categorical(Binary) data

```{r}
wcgs %>% filter(dibpat == "Type B") %>% pull(smoke) %>% table() %>% prop.table() %>% kable()
```

## Classify continuous data into categorical data

```{r}
#reclassify continuous data
aug$classification <- cut(aug$Score, c(0, 8, 17, 24, 52), include.lowest = TRUE)
#rename categories
levels(aug$classification)<-c("None", "Mild", "Moderate", "Severe")
#create table
aug%>%
  filter(Treatment_Group=="Augmentation")%>%
  pull(classification)%>%
  table()%>%
  prop.table()%>%
  kable()
```

## Visualized categorical data by barplot

```{r}
#visualized by barplot
wcgs %>% pull(agec) %>% 
  table() %>% 
  barplot(main = "Distribution of Ages",
          xlab = "Age group",
          ylab = "Number of people")
```


# Test for mean (Continuous data)

## One Sample

```{r}
#one sample ttest
ttest(psa$pretxpsa, null.hypoth = 100)
#one sample ttest (removing NA)
ttest(psa$pretxpsa[!is.na(psa$bss)], null.hypoth = 100)
```

## Two sample

```{r}
#unpaired
sbp_typeA <- wcgs %>% filter(dibpat == "Type A") %>% pull(sbp)
sbp_typeB <- wcgs %>% filter(dibpat == "Type B") %>% pull(sbp)
ttest(sbp_typeA, sbp_typeB)
#paired
ttest(sep$n35R, sep$n35L, matched = TRUE)
```

# Categorize continuouse data

```{r}
#q1 create a new binary variable
carotine$arm_bi<-ifelse(carotine$arm<30, 1, 0)
#1= 30 mg/day or greater of beta-carotene.
carotine$diff<- (carotine$followup.vite-carotine$baseline.vite)
carotine$diff_bi<-ifelse(carotine$diff>=0, 1, 0)
#1= difference ≥ 0
```

```{r}
carotinetable<- table(carotine$arm_bi, carotine$diff_bi)
rownames(carotinetable) <- c(">=30", "<30")
colnames(carotinetable) <- c("difference<0", "difference>=0")
carotinetable
```

# Test for proportion (Binary or Categorical data)

## Check whether the samples are normal approximation

```{r}
psa %>% filter(severe == 1) %>% pull(psaover100) %>% table()
#the rule of thumb: np≥10 and n(1−p)≥10 where n=25 and p=0.5 (this is the value of p under H0). We get 13 in either of case so we can use a one sample z-test for the proportion (the normal approximation)
proptesti(x1 = 17, n1 = 26, null.hypoth = 0.5,conf.level = 0.95,alternative = "two.sided", correct = TRUE)
```

## One Sample 

### Using stimulation

```{r}
#demo
nexperiments <- 1e6 # 1 million experiments
ninfants <- 16 # number of total trials within an experiment
set.seed(2) # set a seed so that our results are replicable
samples <- rbinom(nexperiments, ninfants, prob = 0.5)
hist(samples)
#finding confidence interval
nexperiments <- 1e6 # 1 million experiments
nptps <- 107 # number of total trials within an experiment
set.seed(512)
#p0 = 0.13
samples13 <- rbinom(nexperiments, nptps, prob = 0.13)
quantile(samples13, probs = c(0.025, 0.975))
```

### Exact test

```{r}
#when the x1 and n1 are already known
proptesti(x1 = 14, n1 = 16, exact = TRUE)
#when null is not 0.5, and alternative is greater but not "not equal"
proptesti(x1 = 14, n1 = 16, exact = TRUE, 
          null.hypoth = 0.6, alternative = "greater")
#two sided test
proptesti(x1 = 146, n1 = 191, null.hypoth = 0.66, exact = TRUE,alternative = "two.sided")
#the way to get x1 and n1
mri %>% filter(highldl == 1) %>% nrow()
mri %>% filter(highldl == 1, death5 == 1) %>% nrow()
#one sample z-test for proportion (not applied to rule of thumb)
proptesti(x1 = 14, n1 = 107, exact = TRUE, null.hypoth = 0.16)
#one sample exact binomial for proportion (applied to rule of thumb)
proptesti(x1 = 14, n1 = 107, correct = TRUE, null.hypoth = 0.16)
#more example
a<-chs_recode%>%filter(age>=65 & mi==1)%>%nrow()
b<-chs_recode%>%filter(age>=65)%>%nrow()
proptesti(a,b,null.hypoth = 0.1, exact=FALSE, correct=TRUE)
```

## Two sample

### Large sample size (>=5)

```{r}
#paired
chs_recode_new <- chs_recode[701:1000,] #with only the last 300 patients in the dataset
malemi2<-chs_recode_new%>%filter(sex=="male")%>%pull(mi)
femalemi2<-chs_recode_new%>%filter(sex=="female")%>%pull(mi)
proptest(malemi2, femalemi2)
```

```{r}
#two group Chi-square
data<-matrix(c(86, 8, 19914, 19992), nrow=2)
epi.2by2(data, method="cohort.count")
#more example
epi.2by2(table(carotine$arm_bi, carotine$diff_bi), method="cohort.count")
```

### Small sample size (<5)

```{r}
#recategorized
carotinetable5l<- table(carotine$arm, carotine$diff_bi)
rownames(carotinetable5l) <- c("0", "15", "30", "45", "60")
colnames(carotinetable5l) <- c("difference<0", "difference>=0")
carotinetable5l
```
```{r}
fisher.test(carotinetable5l)
```

### Paired Binary data

```{r}
mat <- matrix(c(195, 5, 15, 35), ncol = 2, nrow = 2)
mcnemar_res <- mcnemar.test(mat)
```


## More than two groups of samples

```{r}
#Chi-square
mat<-matrix(c(89,640,194,3190))
chisq.test(mat)
```


# Correlation

```{r}
cor(wcgs$weight, wcgs$sbp)
cor(wcgs$weight, wcgs$sbp, method = "kendall")
cor(wcgs$weight, wcgs$sbp, method = "spearman")
```

## Fitting the Model

```{r}
model <- regress("mean", sbp ~ weight, data = wcgs)
model
```

## Scatterplot

```{r}
plot(sbp ~ weight, data = wcgs, 
     xlab = "Weight (lbs)", ylab = "Systolic blood pressure (mm Hg)",
     pch = 16, cex = 0.6)
abline(model, col = "blue", lwd=2)
```

